Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-21 manuscript/response_to_reviews.Rmd

L266-267: Don’t think you can test the effects of crown dominance going back this far. How can you be sure a tree was dominant/suppressed back in 1966?

**We have dropped crown position from the model, and now use it only to give a picture of how microenvironmental conditions vary with tree height under contemporary conditions (Fig. 2).**

L287: specify how they were related to drought. Positive? Negative?

**This has been revised to read, "In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Fig. **4**; Tables 1,S4-S7)."**

@@

**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al. (2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms. "**

Secondly, the choice of traits does not appear well justified, as it seems to be driven more by easiness of measurement than by relevance, and it mixes different types of hypotheses operating at different levels. On one hand there is tree height and crown position, which are measured at the level of individual trees and correspond to closely related hypotheses. On the other hand there are ring-porosity, wood density, LMA and water relations traits, which are measured at the species level and seem to be totally disconnected from the hypotheses regarding tree height. In addition, it is very hard to compare the explanatory power of these variables, as done in the paper, as they are based on very different sample sizes (571 trees vs 12 species) and respond to different sources of variability. In fact the way height effects are analyzed mixes the within- and between-species levels, which makes it difficult to interpret the results. Finally, I think it would be more correct to refer to PLA and TLP as water relations traits rather than hydraulic traits.

**We respond to this one point at a time:**

*the choice of traits does not appear well justified, as it seems to be driven more by easiness of measurement than by relevance*

**Ease of measurement is part of what makes $PLA_{dry}$ and $\pi_{tlp}$ interesting. As we now clarify in the introduction:**

**"However, [traits such as P50, P88, and hydraulic safety margin] are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016].*The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.*"**

*it mixes different types of hypotheses operating at different levels. On one hand there is tree height and crown position, which are measured at the level of individual trees and correspond to closely related hypotheses. On the other hand there are ring-porosity, wood density, LMA and water relations traits, which are measured at the species level and seem to be totally disconnected from the hypotheses regarding tree height. In addition, it is very hard to compare the explanatory power of these variables, as done in the paper, as they are based on very different sample sizes (571 trees vs 12 species) and respond to different sources of variability. In fact the way height effects are analyzed mixes the within- and between-species levels, which makes it difficult to interpret the results.*

**There is nothing fundamentally wrong with a model comparing individual- and species- level effects, but we understand why it was confusing and believe that the revised presentation of the analysis should make all of this easier to follow. Improvements related to this comment include: (1) revision to how traits are selected for inclusion as candidate variables for the full model, which provides a more appropriate test of individual traits and guards against over-fitting; (2) moving of former tables 4 and 5 to the SI, and replacing fig. 5 with a figure visualizing the model results; (3) adding a couple new figures showing variation in $Rt$ (Fig. 3) and predictor variables (Fig. S3) across species; and (4) dropping canopy position from the model for predicting $Rt$.

*Finally, I think it would be more correct to refer to PLA and TLP as water relations traits rather than hydraulic traits.*

*RESPONSE NEEDED HERE!*

@@

Thirdly, a more thorough analysis of the distribution of the study species in the sample plot would be required. The paper assumes that all species were subjected to the same drought levels. However, this is unlikely to be true, as different species are likely to occupy different microhabitats in the plot, and hence be subjected to different drought conditions. The authors account for this, to some extent, by including individual-level TWI in some models, but a more thorough analysis of the distribution of the study species as a function of TWI and other relevant micro-environmental variables would be needed to fully interpret the results.

**We have added a new figure to the SI showing variation in predictor variables across species (Fig. S3). While there was some variation across species in $H$ and $TWI$, there were few pronounced differences across species. We now discuss the case of the most anomalous species, *Fagus grandifolia*, in the discussion section. We also determined that there is little systematic variation in $PLA_{dry}$ and $\pi_{tlp}$ with respect to $H$ or $TWI$ (analysis not shown). This, together with the fact that our statistical models do account for $TWI$, should minimize bias related to distribution of the study species within the plot**

@@

Minor or more technical suggestions:

L59-79: the arguments presented in this paragraph mix different scales, from within-individual to within- or among-communities. I think the arguments could be better structured to reflect the sources of variability involved and, ideally, focus on the scale that will be addressed in the paper.

**We have revised this paragraph to provide some more specifics on the scale and setting of these tests.**

@@

L67: I suggest toning down the sentence to something like ‘Wider xylem conduits are associated to higher vulnerability to embolism during drought’, as the mechanism linking wider conduits with embolism vulnerability is far from clear and probably indirect.

**We have revised this to read, "Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018]..."**

@@

L81: you seem to use drought resistance and drought tolerance interchangeably, which may be confusing. I think it is important to clarify these definitions early on in the paper.

*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/124*

@@

L97: ‘greater variation’ that what?

**This was removed. The text here now reads, "More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; *Zhu et al. 2018*]. Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; *Zhu et al. 2018*; but see *Farrell et al. 2017*]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested.  "**

@@

L100-101: that is not true really. See for example Zhu et al. (2018) in Tree Phys, Farrell et al. (2017) in PCE…

**We have added citations to both these papers and revised the sentence so that it is true: "The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested."**

@@

L160: you mean a log-log regression, right?

**Yes. fixed.**

L181: N=3 individuals per species is really quite low. In addition, by sampling up to 8 m height you would probably be sampling less exposed branches from taller trees, which could confound your results.

**More samples would be ideal, but was not feasible given the time and resources available, and have no way to sample more now (in part because of COVID-19). We note that intraspecific SD was generally low with respect to mean values and interspecific variation, and that a sample size of n=3 is not unprecendented for this type of analysis [@medeiros_extensive_2019]. We have added a figure to the SI (Fig. S3) showing variation in traits across species.**

*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/125*

### 2020-07-21 manuscript/climate_sensitivity_SI.Rmd
update S3
worked on R4 comments

\newpage
![Figure S3: **Distribution of independent variables by species.** Species that are assigned the same letter are not significantly different from each other with regard to the tested variable. ](tables_figures/publication/independent_variables_by_species.png)

\newpage
![Figure S4: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)

\newpage
![Figure S5: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)

>>>>

\newpage

![Figure S3: **Distribution of independent variables by species.** Species that are assigned the same letter are not significantly different from each other with regard to the tested variable. ](tables_figures/publication/FigureS3_traits_signif.png)

\newpage

![Figure S4: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)

\newpage

![Figure S5: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
finished R3!
worked on R4 comments

Debates have also arisen regarding the traits influencing tree growth responses to drought. Studies in temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al. 2007, Olson et al. 2020*) and does not resolve differences among the many species within each category. Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent. For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].

In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance [@anderegg_hydraulic_2018]. 
However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
These include leaf area shrinkage upon desiccation ($PLA_{dry}$) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. 
Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016].
The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

@@

Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 and 2019 (n=1,518 trees). Methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018]. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H] \sim ln[DBH]$). For species with insufficient height data to create reliable species-specific allometries (n=2, JUNI and FRAM), heights were calculated from an equation developed by combining the height measurements across all species. 
We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$.

@@

In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were linked to drought responses (Fig. **4**; Tables 1,S4-S7). 
Both had consistent signs across all droughts, and their inclusion at least marginally improved the model ($\Delta$AICc > 1.0) for at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model. 
$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
It was also included in some of the top models for 1999 (Tables S6-S7).

@@

Concerted measurement of tree-rings and leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Fig. **4**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019], and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional type. Further, it adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017]. Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests. The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts ([Christoffersen et al. 2016](https://pearl.plymouth.ac.uk/handle/10026.1/12918)) [@powell_differences_2017].

>>>>

Debates have also arisen regarding the traits influencing tree growth responses to drought. Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al. 2007, Olson et al. 2020*) and does not resolve differences among the many species within each category. Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance [e.g., in a tropical rainforest; @marechaux_leaf_2019], or the direction of response was not always consistent. For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].

In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests (*[Anderegg et al. 2016](www.pnas.org/cgi/doi/10.1073/pnas.1525678113)*). 
However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; *Zhu et al. 2018*]. 
Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; *Zhu et al. 2018*; but see *Farrell et al. 2017*].
The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested.

@@

Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 and 2019 (n=1,518 trees). Methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018]. Species-specific height allometries were developed (Table S2) using log-log regression ($ln[H] \sim ln[DBH]$). For species with insufficient height data to create reliable species-specific allometries (n=2, JUNI and FRAM), heights were calculated from an equation developed by combining the height measurements across all species. 
We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$.

@@

In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Fig. **4**; Tables 1,S4-S7). 
Both had consistent signs across all droughts, and their inclusion at least marginally improved the model ($\Delta$AICc > 1.0) for at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model. 
$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
It was also included in some of the top models for 1999 (Tables S6-S7).

@@

Concerted measurement of tree-rings and leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Fig. **4**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019; *Zhu et al. 2018*], and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional type. Further, it adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017]. Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests. The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts ([Christoffersen et al. 2016](https://pearl.plymouth.ac.uk/handle/10026.1/12918)) [@powell_differences_2017].
